---
title: "Pertussis Challenge 3.2_CCL+KAT+STAT"
author: "Runqi Zhang"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    number_sections: true
    keep_tex: true
---

```{r set up}
# Load libraries and suppress unnecessary output
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(agua)          # H2O AutoML
  library(edgeR)         # For TMM normalization
  library(GSVA)          # Gene set variation analysis
  library(biomaRt)       # Gene ID mapping
  library(glmnet)        # Regression
  library(knitr)         # Output formatting
  library(tibble)        # For handling row names
  library(impute)
})

# Define working directory and initialize H2O
workDir <- "C:/Users/zhang/Desktop/cmi-pb-3rd-final/Runqi/CMI-PB"
options(readr.show_col_types = FALSE)
agua::h2o_start()

```

```{r read input}
# Function to read input data for a given year
read_data <- function(year, type = "LD") {
  list(
    pts = read_tsv(file.path(workDir, paste0("data/", year, type, "_subject.tsv"))),
    sample = read_tsv(file.path(workDir, paste0("data/", year, type, "_specimen.tsv"))),
    ge = read_tsv(file.path(workDir, paste0("data/", year, type, "_pbmc_gene_expression.tsv")))
  )
}

# Load datasets for 2020–2023
data2020 <- read_data("2020")
data2021 <- read_data("2021")
data2022 <- read_data("2022")
data2023 <- read_data("2023", type = "BD")

```

```{r prepare outcome}
# Function to prepare fold change (FC) outcomes
prepare_outcome <- function(ge_data, sample_data, pts_data, gene_id, days = c(0, 3)) {
  ge_data %>%
    filter(versioned_ensembl_gene_id == gene_id) %>%
    inner_join(sample_data, by = "specimen_id") %>%
    inner_join(pts_data, by = "subject_id") %>%
    filter(planned_day_relative_to_boost %in% days) %>%
    dplyr::select(subject_id, planned_day_relative_to_boost, tpm) %>%
    pivot_wider(names_from = planned_day_relative_to_boost, values_from = tpm, names_prefix = "tpm_") %>%
    mutate(tpm_FC = scale(tpm_3 / tpm_0)) # Calculate FC and scale
}

# Prepare outcome data for all years
yDF <- prepare_outcome(data2020$ge, data2020$sample, data2020$pts, "ENSG00000277632.1")
y2DF <- prepare_outcome(data2021$ge, data2021$sample, data2021$pts, "ENSG00000277632.1")
y3DF <- prepare_outcome(data2022$ge, data2022$sample, data2022$pts, "ENSG00000277632.1")

```

#top 20 CCL paralogs
#CCL3 ENSG00000277632.1
#CCL3L3 ENSG00000276085.1
#CCL4 ENSG00000275302.1
#CCL4L2 ENSG00000276070.4
#CCL5 ENSG00000271503.5
#CCL14 ENSG00000276409.4
#CCL15 ENSG00000275718.1
#CCL18 ENSG00000275385.1
#CCL22 ENSG00000102962.4
#CCL23 ENSG00000274736.4
#CCL26 ENSG00000006606.8
#CCL24 ENSG00000106178.6
#CCL16 ENSG00000275152.4
#CCL1 ENSG00000108702.1
#CCL17 ENSG00000102970.10
#CCL25 ENSG00000131142.14
#CCL7 ENSG00000108688.11
#CCL19 ENSG00000172724.11
#CCL8 ENSG00000108700.4
#XCL1 ENSG00000143184.4
#XCL2 ENSG00000143185.3

#KAT -> STAT1 -> CCL3 pathway
#KAT7 ENSG00000136504.11
#PHF10 ENSG00000130024.14
#RSF1 ENSG00000048649.14
#KAT6B ENSG00000156650.12
#DPF1 ENSG00000011332.19
#KAT5 ENSG00000172977.12
#KAT8 ENSG00000103510.19
#DPF3 ENSG00000205683.11
#KAT6A ENSG00000083168.9
#DPF2 ENSG00000133884.9

#STAT1 ENSG00000115415.18
#STAT2 ENSG00000170581.13
#STAT3 ENSG00000168610.14
#STAT4 ENSG00000138378.17
#STAT5A ENSG00000126561.16
#STAT6 ENSG00000166888.11
#STAT5B ENSG00000173757.9

```{r gene expression}
# Target genes
target_genes <- c(
  "ENSG00000277632.1", # CCL3
  "ENSG00000276085.1", # CCL3L3
  "ENSG00000275302.1", # CCL4
  "ENSG00000276070.4", # CCL4L2
  "ENSG00000271503.5", # CCL5
  "ENSG00000276409.4", # CCL14
  "ENSG00000275718.1", # CCL15
  "ENSG00000275385.1", # CCL18
  "ENSG00000102962.4", # CCL22
  "ENSG00000274736.4", # CCL23
  "ENSG00000006606.8", #CCL26
  "ENSG00000106178.6", #CCL24 
  "ENSG00000275152.4", #CCL16
  "ENSG00000108702.1", #CCL1
  "ENSG00000102970.10", #CCL17
  "ENSG00000131142.14", #CCL25
  "ENSG00000108688.11", #CCL7
  "ENSG00000172724.11", #CCL19
  "ENSG00000108700.4", #CCL8
  "ENSG00000143184.4", #XCL1
  "ENSG00000143185.3", #XCL2
  "ENSG00000136504.11", #KAT7 
  "ENSG00000130024.14", #PHF10
  "ENSG00000048649.14", #RSF1
  "ENSG00000156650.12", #KAT6B
  "ENSG00000011332.19", #DPF1
  "ENSG00000172977.12", #KAT5
  "ENSG00000103510.19", #KAT8
  "ENSG00000205683.11", #DPF3
  "ENSG00000083168.9", #KAT6A
  "ENSG00000133884.9", #DPF2
  "ENSG00000115415.18", #STAT1
  "ENSG00000170581.13", #STAT2
  "ENSG00000168610.14", #STAT3
  "ENSG00000138378.17", #STAT4
  "ENSG00000126561.16", #STAT5A
  "ENSG00000166888.11", #STAT6
  "ENSG00000173757.9" #STAT5B
)

# Function to extract day 0 TPM values
extract_day0_tpm <- function(ge_data, sample_data, pts_data, target_genes) {
  ge_data %>%
    filter(versioned_ensembl_gene_id %in% target_genes) %>%
    inner_join(sample_data, by = "specimen_id") %>%
    inner_join(pts_data, by = "subject_id") %>%
    filter(planned_day_relative_to_boost == 0) %>%
    dplyr::select(subject_id, versioned_ensembl_gene_id, tpm) %>%
    pivot_wider(
      names_from = versioned_ensembl_gene_id,
      values_from = tpm,
      names_prefix = "tpm_"
    ) %>%
    replace(is.na(.), 0) %>% # Replace NA with 0
    column_to_rownames("subject_id") # Use subject_id as rownames
}

# Extract day 0 TPM values for each dataset
xDF <- extract_day0_tpm(data2020$ge, data2020$sample, data2020$pts, target_genes)
x2DF <- extract_day0_tpm(data2021$ge, data2021$sample, data2021$pts, target_genes)
x3DF <- extract_day0_tpm(data2022$ge, data2022$sample, data2022$pts, target_genes)
x4DF <- extract_day0_tpm(data2023$ge, data2023$sample, data2023$pts, target_genes)

# Ensure consistent columns across datasets
common_cols <- Reduce(intersect, list(colnames(xDF), colnames(x2DF), colnames(x3DF), colnames(x4DF)))
xDF <- xDF[, common_cols]
x2DF <- x2DF[, common_cols]
x3DF <- x3DF[, common_cols]
x4DF <- x4DF[, common_cols]

```

```{r training&predict}
# Train model with H2O AutoML
trainDF <- rbind(xDF, x2DF, x3DF) %>%
  as.data.frame() %>%
  mutate(tpm_FC = c(yDF$tpm_FC, y2DF$tpm_FC, y3DF$tpm_FC))

trainDF$tpm_FC <- (log1p(trainDF$tpm_FC))

#Applyk-NNimputation
 train_matrix<-as.matrix(trainDF) #Converttomatrix
 imputed_matrix<-impute.knn(train_matrix)$data
 trainDF<-as.data.frame(imputed_matrix)#Convertbacktodataframe
 
set.seed(3)
auto_fit <- auto_ml() %>%
  set_engine("h2o", max_runtime_secs = 5) %>%
  set_mode("regression") %>%
  fit(tpm_FC ~ ., data = trainDF)
```

```{r validation}
# Predict on training data
train_predictions <- predict(auto_fit, new_data = trainDF)$.pred
 
# Calculate correlations
pearson_cor <- cor(train_predictions, trainDF$tpm_FC, method = "pearson")
spearman_cor <- cor(train_predictions, trainDF$tpm_FC, method = "spearman")

# Display correlation results
cat("Pearson Correlation: ", pearson_cor, "/n")
cat("Spearman Correlation: ", spearman_cor, "/n")

# Create a correlation plot
library(ggplot2)

correlation_plot <- ggplot(data = data.frame(
  Predicted = train_predictions,
  Actual = trainDF$tpm_FC
), aes(x = Predicted, y = Actual)) +
  geom_point(alpha = 0.6, color = "blue") +  # Scatter plot
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
  labs(
    title = "Model Validation: Predicted vs Actual TPM_FC",
    subtitle = paste0("Pearson: ", round(pearson_cor, 2), 
                      " | Spearman: ", round(spearman_cor, 2)),
    x = "Predicted TPM_FC",
    y = "Actual TPM_FC"
  ) +
  theme_minimal()

# Display the plot
print(correlation_plot)

```

```{r predict and rank}
# Predict and rank
yhat <- predict(auto_fit, new_data = x4DF)$.pred
rhat <- rank(-1 * yhat, ties.method = "first")
print(cbind(rownames(x4DF), rhat))

```

```{r update submission file}
# Update submission file with rankings
submission_file <- file.path(workDir, "3rdChallengeSubmissionTemplate_revised.tsv")
data <- read_tsv(submission_file)

ranking_df <- data.frame(
  SubjectID = as.numeric(rownames(x4DF)),
  "3.2) CCL3-D3-FC-Rank" = rhat,
  check.names = FALSE
)

data <- data %>%
  mutate(
    `3.2) CCL3-D3-FC-Rank` = ifelse(
      SubjectID %in% ranking_df$SubjectID,
      ranking_df$`3.2) CCL3-D3-FC-Rank`[match(SubjectID, ranking_df$SubjectID)],
      `3.2) CCL3-D3-FC-Rank`
    )
  )

write_tsv(data, submission_file)

```


```{r session info}
# End H2O session and print session info
agua::h2o_end()
sessionInfo()

```

